---
output:
  pdf_document: default
  html_document: default
---
```{r, echo=FALSE}
#| echo: false
#| warning: false
#| message: false

library(forecast)
library(tseries)
library(vars)
library(ggplot2)
library(gridExtra)
library(tidyverse)
library(stats)
library(lubridate)
library(fGarch)
library(rugarch)
library(knitr)
library(MTS)
cocoa_data <-read_csv(here::here("data/cleaned_data/cocoa_data.csv"))
```

# Data
```{r, echo=FALSE}
#| echo: false
#| include: false
#| warning: false
#| message: false

cocoa_data$YearMonth <- as.Date(paste0(cocoa_data$YearMonth, "-01"))
cocoa_data$Price_Lag1 <- c(NA, head(cocoa_data$Price, -1))
cocoa_data$Price_Lag2 <- c(NA, NA, head(cocoa_data$Price, -2))
cocoa_data <- na.omit(cocoa_data)
var_data <- cocoa_data[, c("Price", "Price_Lag1", "Price_Lag2")]

var_data$Price_Diff <- c(NA, diff(var_data$Price))
adf.test(na.omit(var_data$Price_Diff), k = 7)

var_data <- na.omit(var_data)
VARselect(var_data$Price_Diff, lag.max = 7, type = "const")
```

```{r, echo=FALSE}
#| include: false
#| echo: false
#| warning: false
#| message: false

var_data_clean <- na.omit(cocoa_data[, c("Price", "Price_Lag1", "Price_Lag2", "Monthly_Avg_Temp", "Monthly_Avg_Precip", "Exchange_Rate")])
adf.test(var_data_clean$Price)
adf.test(var_data_clean$Price_Lag1)
adf.test(var_data_clean$Price_Lag2)
adf.test(var_data_clean$Monthly_Avg_Temp)
adf.test(var_data_clean$Monthly_Avg_Precip)
adf.test(var_data_clean$Exchange_Rate)

# Apply differencing and add NA for first value
var_data_clean$Price_Diff <- c(NA, diff(var_data_clean$Price, differences = 1))
var_data_clean$Price_Lag1_Diff <- c(NA, diff(var_data_clean$Price_Lag1, differences = 1))
var_data_clean$Price_Lag2_Diff <- c(NA, diff(var_data_clean$Price_Lag2, differences = 1))
var_data_clean$ExRate_Diff <- c(NA, diff(var_data_clean$Exchange_Rate, differences = 1))

# Remove rows with NA values (i.e., the first row for each differenced column)
var_data_clean <- na.omit(var_data_clean[, c("Price_Diff", "Price_Lag1_Diff", "Price_Lag2_Diff", "Monthly_Avg_Temp", "Monthly_Avg_Precip", "ExRate_Diff")])
#var_data_clean$Price_Lag1_Diff <- diff(var_data_clean$Price_Lag1, differences = 1)


# Perform ADF test on the differenced variables
adf.test(var_data_clean$Price_Diff)
adf.test(var_data_clean$Price_Lag1_Diff)
adf.test(var_data_clean$Price_Lag2_Diff)
adf.test(var_data_clean$ExRate_Diff)

```

We now consider the VARMA model due to the volatility of cocoa prices, while still incorporating parameters from the ARMA model. To ensure the variables are strong significant (p-value < 0.05), we first apply set difference on the variables that are not statistically significant (p-value > 0.05), such as `cocoa_price`, `price_lag`, and `currency_exchange_rate`. With these significant variables identified, we proceed to fit the VARMA model by selecting an appropriate lag order.


```{r, echo=FALSE}
#| include: false
#| echo: false
#| warning: false
#| message: false

var_model <- VAR(var_data_clean[, c("Price_Diff", "Price_Lag1_Diff", "Price_Lag2_Diff", "ExRate_Diff", "Monthly_Avg_Temp", "Monthly_Avg_Precip")], p = 1)
lag_select <- VARselect(var_data_clean[, c("Price_Diff", "Price_Lag1_Diff", "Price_Lag2_Diff", "ExRate_Diff", "Monthly_Avg_Temp", "Monthly_Avg_Precip")], lag.max = 10)
```


```{r, echo=FALSE}
lag_table <- data.frame(
  names(lag_select$selection),
  as.numeric(lag_select$selection)
)
kable(lag_table, col.names = NULL, caption = "Lag Order Selection")
```

From the results, we will consider lag=1, 2, 4. We will use model selection to decide which one is better.

```{r, echo=FALSE}
#| include: false
#| echo: false
#| warning: false
#| message: false

varma_model <- VARMA(
  var_data_clean[, c("Price_Diff", "ExRate_Diff", "Monthly_Avg_Temp")], 
  p = 1, q = 2)
varma_model_lag2 <- VARMA(
  var_data_clean[, c("Price_Diff", "ExRate_Diff", "Monthly_Avg_Temp")], 
  p = 2, q = 2)
varma_model_lag4 <- VARMA(
  var_data_clean[, c("Price_Diff", "ExRate_Diff", "Monthly_Avg_Temp")], 
  p = 4, q = 2)
```

```{r, echo=FALSE}
# Extract AIC and BIC values
compute_AIC <- function(model) {
  # Number of estimated parameters
  k <- length(unlist(model$coef))  # Total estimated parameters
  # Residual sum of squares (approximation for log-likelihood)
  sigma2 <- sum(model$residuals^2, na.rm = TRUE) / length(model$residuals)
  # Approximate log-likelihood
  logL <- -0.5 * length(model$residuals) * log(sigma2)
  # Compute AIC
  AIC_value <- 2 * k - 2 * logL
  return(AIC_value)
}
aics <- c(compute_AIC(varma_model), compute_AIC(varma_model_lag2), compute_AIC(varma_model_lag4))
aic_results <- data.frame(Model = c("VARMA(p=1, q=2)", "VARMA(p=2, q=2)", "VARMA(p=4, q=2)"),
                          AIC = aics)
kable(aic_results, caption = "AIC for Different Lag Orders")


```

We built three VARMA models with different lag orders (p = 1, 2, 4, q = 2) to forecast price fluctuations. The difference between these models lies in the number of past observations (lags) they consider for predicting future values. To compare them, we used the AIC, which balances model fit and complexity. The results show that VARMA(p=1, q=2) has the lowest AIC, making it the best model among the three. Below is the model expression

```{r, echo=FALSE}
#| include: false
#| echo: false
#| warning: false
#| message: false

varma_model <- VARMA(
  var_data_clean[, c("Price_Diff", "ExRate_Diff", "Monthly_Avg_Temp")], 
  p = 1, q = 2) # for AR and MA respectively

summary(varma_model)
```

```{=tex}
\begin{align*} 
\text{PriceDiff}_{t} = &\ \beta_0 + \beta_1 \text{PriceDiff}_{t-1} + \beta_2 \text{ExRateDiff}_{t-1} + \beta_3 \text{MonthlyAvgTemp}_{t-1} + \\
&\ \beta_4 \epsilon_{\text{PriceDiff}, t-1} + \beta_5 \epsilon_{\text{ExRateDiff}, t-1} + \beta_6 \epsilon_{\text{MonthlyAvgTemp}, t-1} + \\
&\ \beta_7 \epsilon_{\text{PriceDiff}, t-2} + \beta_8 \epsilon_{\text{ExRateDiff}, t-2} + \beta_9 \epsilon_{\text{MonthlyAvgTemp}, t-2} + \epsilon_t
\end{align*}

```

Where:

-   `p = 1` means the model uses the lagged values from the previous time step (t-1) for PriceDiff, ExRateDiff, and MonthlyAvgTemp (AR part)

-   `q = 2`: means the model uses lagged residuals (errors) from the previous two time steps (t-1 and t-2) for PriceDiff, ExRateDiff, and MonthlyAvgTemp (MA part).

-   $\beta_1, \beta_2, ...$ are the coefficients for the lagged variables and error terms
$\epsilon_t$ represents the residual (error) term at time t.

```{r, echo=FALSE}
#| label: fig-model-check
#| fig-cap: "The residuals in the VARMA(1, 2) model shows that most of the bars, which represent autocorrelation at different lags, fall within the blue confidence interval bands, typically the default for ACF plots. This indicates that the residuals do not exhibit significant autocorrelation and are random. Therefore, the residuals can be considered white noise, suggesting that the model has effectively captured the underlying patterns. Based on this, we will forecast the price fluctuations over time using VARMA(1, 2) model."
#| echo: false
#| warning: false
#| message: false

acf(residuals(varma_model))
```

\newpage

```{r, echo=FALSE}
#| include: false
#| echo: false
#| warning: false
#| message: false

# Prepare the data for VAR model (ensure it's a time series object)
ts_data <- ts(var_data_clean, frequency = 12)  # Use your actual data and adjust the frequency if necessary
varma_forecast <- VARMApred(varma_model, h = 3)
#forecast_lower <- varma_forecast$pred - 1.96 * varma_forecast$se.err
#forecast_upper <- varma_forecast$pred + 1.96 * varma_forecast$se.err
# Time points for the forecast
forecast_time <- seq(from = length(var_data_clean$Price_Diff) + 1, by = 1, length.out = length(varma_forecast$pred))
```

```{r, echo=FALSE}
#| label: fig-varma-forecast
#| fig-cap: "The forecast for price changes indicates significant short-term volatility, with an initial sharp increase followed by fluctuations before stabilizing around a moderate positive value. The first two steps show a large jump (+88.74 to +241.41), followed by a drop (-20.06), and then a gradual stabilization around 20–30. This suggests that prices may experience a sharp initial surge before settling into a more predictable upward trend. If applied to actual prices, this pattern implies a high-risk, high-reward scenario in the short term, followed by steady growth. For trading or investment decisions, this means navigating an initial period of uncertainty before benefiting from more stable price movements."
#| echo: false
#| warning: false
#| message: false

plot(forecast_time, varma_forecast$pred, type = "o", col = "blue",
     xlab = "Months Ahead", ylab = "Price Fluctuation",
     main = "VARMA Forecast of Price Fluctuation for the Next 5 Months")

```

```{r, echo=FALSE}
#| include: false
#| echo: false
#| warning: false
#| message: false

# Generate forecasts and predicted values
varma_forecast <- VARMApred(varma_model, h = 10)  # Forecast for the next 10 months
pred_values <- varma_forecast$pred
# Assume we have actual values for comparison (replace with actual data)
actual_values <- var_data_clean$Price_Diff[(nrow(var_data_clean)-9):nrow(var_data_clean)]
```


```{r, echo=FALSE}
# Compute error metrics
rmse <- sqrt(mean((actual_values - pred_values)^2, na.rm = TRUE))
mae <- mean(abs(actual_values - pred_values), na.rm = TRUE)
mape <- mean(abs((actual_values - pred_values) / actual_values) * 100, na.rm = TRUE)

# Create a table with results
accuracy_table <- data.frame(
  Metric = c("RMSE", "MAE", "MAPE"),
  Value = c(rmse, mae, mape)
)
kable(accuracy_table, caption = "VARMA Model Accuracy Metrics")
```

The table presents the accuracy metrics for the VARMA model. The RMSE of 1331.574 and MAE of 1097.820 indicate the average magnitude of errors between predicted and actual values, with lower values reflecting better model performance.

