---
title: "What to Expect with Cocoa Futures in Ghana?"
subtitle: "Forecasting Cocoa Price: A Time Series Approach to Commodity Price Modeling"
author: 
  - Emily Pan, Manjun Zhu, Edan Wong
format:
  pdf:
    pdf_engine: xelatex
    fontsize: 10pt
    mainfront: Times New Roman
bibliography: references.bib
date: April 4, 2025
date-format: long
number-sections: true
geometry: margin=0.75in
editor: 
  markdown: 
    wrap: 72
---

# Introduction

Cocoa is a major agricultural commodity in Ghana. The industry employs
over 850,000 farmers and generates \$2 billion in foreign exchange
annually [@GhanaCocoaBoard]. Cocoa futures are traded primarily on the
Intercontinental Exchange (ICE), which plays a significant role in
global pricing, market expectations, and risk management. Although Ghana
does not have a domestic cocoa futures market, its daily cocoa prices,
export contracts, and farmgate prices are affected by international
cocoa futures benchmarks [@Nunley2024]. Daily cocoa prices show the
current market value for immediate delivery, while futures prices
indicate expected future values based on supply, demand, and market
predictions influenced by macroeconomic factors and weather conditions.

Recent developments in the global cocoa market have highlighted the
volatility and sensitivity of cocoa prices to both environmental shocks
and market speculation. From 1994 to 2023, cocoa prices showed long-term
growth, rising from approximately US\$1,400 per ton to around US\$4,000
per ton [@InternationalCocoaOrganization]. However, in 2024, cocoa
prices experienced unprecedented fluctuations—peaking at US\$12,072 per
ton in February before sharply declining by September
([@GhanaCocoaBoard, @InternationalCocoaOrganization]. These fluctuations
have been driven by supply constraints, rising demand, and adverse
production conditions linked to climate change, aging trees, pests, and
disease.

In response, the Ghana Cocoa Board raised the producer price to GH¢3,100
per 64kg bag to help farmers benefit from high international prices and
cushion the effects of market volatility [@GhanaCocoaBoard]. The ongoing
threats related to climate change highlight the urgent need for
comprehensive and long-term strategies aimed at improving forecasting,
stabilizing prices, and fostering sustainable practices in cocoa
agriculture.

This study aims to improve forecasting accuracy for agricultural
commodity prices, particularly cocoa, using time series models such as
ARIMA, ARIMAX, GARCH, and VARMA, alongside machine learning approaches
like XGBoost. By integrating exogenous variables and modelling price
levels and volatility, we seek to offer a comprehensive framework for
understanding and anticipating market dynamics.

# Literature Review

The ARIMA model is a foundational tool in time series forecasting but
often struggles in volatile markets. For instance, Wang, Yue, Wei, and
Lv [@Wang2017] observed that ARIMA consistently underperformed in
forecasting futures prices for wheat, corn, soybean, and sugar due to
its assumptions of linearity and constant variance, resulting in the
highest forecasting errors among tested models.

To better capture volatility, Katsiampa [@Katsiampa2017] applied
GARCH-type models to Bitcoin and found the AR-CGARCH model most
effective in modeling conditional variance by incorporating both short-
and long-term volatility components. While the study focused on
cryptocurrency, the findings are applicable to cocoa markets, which
similarly exhibit high volatility and are influenced by speculative
behavior and external shocks.

For multivariate relationships, models such as Vector Autoregression
(VAR) and Vector Autoregressive Moving Average (VARMA) are used to
assess the joint impact of exogenous factors like exchange rates and
weather. Mayr and Ulbricht [@Mayr2015], however, cautioned that
log-transforming data in VAR frameworks does not consistently improve
forecast accuracy and recommended testing for stationarity and variance
instability before transformation—an important consideration when
modeling cocoa prices.

Recent literature has increasingly turned to machine learning and deep
learning for more robust forecasting. Menéndez-García, García-Nieto,
García-Gonzalo, and Sánchez Lasheras [@MenendezGarcia2024] compared
ARIMA and VARMA with algorithms such as Support Vector Machines (SVM),
Multivariate Adaptive Regression Splines (MARS), and Multilayer
Perceptrons (MLPs) for forecasting platinum prices. MLP, an artificial
neural network (ANN), achieved the highest accuracy and lowest RMSE,
while ARIMA again performed the worst, reinforcing its difficulty in
capturing non-linear and noisy patterns.

Further supporting the shift toward advanced techniques, Nayak, Alam,
Singh, and Sinha [@Nayak2024] evaluated agricultural price forecasting
models for tomatoes, onions, and potatoes in India. Their research found
that deep learning models like TransformerX and NBEATSX, especially when
incorporating exogenous variables such as weather, consistently
outperformed traditional models including ARIMAX, multiple linear
regression, and several ML algorithms (Random Forest, XGBoost, ANN,
etc.) across various accuracy metrics.

Given the limitations of traditional models in volatile agricultural
markets, hybrid approaches have gained traction. These approaches
combine volatility modeling techniques such as GARCH with the
flexibility and pattern-recognition strengths of ML and DL methods like
XGBoost. As researchers increasingly address challenges like
non-stationarity, seasonality, and noisy data, these integrated models
provide a more robust framework for forecasting prices in markets like
cocoa, where production uncertainty and speculative forces remain
prominent.

# Methodology

This section details the methodologies applied in forecasting cocoa
prices using five different modeling approaches: ARIMA, ARIMAX, GARCH,
VARMA and XGBoost. These models were chosen because of their capability
in capturing both statistical and structural characteristics of the
cocoa price series, including autocorrelation, volatility, multivariate
influence and nonlinear patterns. For each fitted model, its performance
will be evaluated through out-of-sample forecasts using metrics such as
Mean Absolute Error (MAE), Root Mean Squared Error (RMSE) and Mean
Absolute Percentage Error (MAPE).

Initially, the analysis was conducted using weekly data to closely match
the frequency of international cocoa price fluctuations. However, this
approach presented several limitations. A significant proportion of
external datasets, including key variables like exchange rates, cocoa
production figures, and macroeconomic indicators, were only available at
a monthly or lower frequency.

The ARIMA model serves as a baseline univariate model and it will be
first applied to forecasting the weekly cocoa price series after testing
for stationarity using the ADF test because the original cocoa daily
price excluded the weekends’ data and . Where necessary, differencing or
box cox transformation will be used to transform the series to
stationary. The optimal parameters (p, d, q) were identified by
examining the ACF and PACF of the transformed price series and AIC and
BIC will be used to select the best ARIMA model after performing the
residual diagnostic for each ARIMA models, where the diagnostic confirms
the absence of autocorrelation in residuals.

## ARIMA Models

The ARIMA model is used to create a baseline forecast of weekly cocoa
prices. It relies only on past prices and patterns in the data, without
using any external factors. This helps us understand how well the price
can be predicted using just its past values.

```{=latex}
\[
Y_t = \phi_1 Y_{t-1} + \cdots + \phi_p Y_{t-p} + \theta_1 \varepsilon_{t-1} + \cdots + \theta_q \varepsilon_{t-q} + \varepsilon_t
\]

where:

\begin{itemize}
  \item $Y_t$ is the log-transformed weekly cocoa price at time $t$,
  \item $\varepsilon_t$ is a white noise error term,
  \item $p$, $d$, and $q$ are the orders of autoregression, differencing, and moving average, respectively.
\end{itemize}
```
To incorporate environmental drivers of cocoa price, we extended the
univariate ARIMA model into an ARIMAX framework by including exogenous
variables such as weekly precipitation and weekly average temperature,
which helps to test whether including these weather factors improves the
accuracy of cocoa price forecasts compared to using price history alone.
This approach allowed us to assess the linear effect of relevant
predictors while maintaining the ARIMA structure to model
autocorrelation in the residuals.

## ARIMAX Model

The ARIMAX model builds on ARIMA by adding climate variables like
precipitation and temperature. It helps us test whether including these
weather factors improves the accuracy of cocoa price forecasts compared
to using price history alone.

```{=latex}
To incorporate climate-related external influences, the ARIMA model is extended to an ARIMAX($p,d,q$) form:

\[
Y_t = \phi_1 Y_{t-1} + \cdots + \phi_p Y_{t-p} + \theta_1 \varepsilon_{t-1} + \cdots + \theta_q \varepsilon_{t-q} + \beta_1 X_{1t} + \cdots + \beta_k X_{kt} + \varepsilon_t
\]

where:

\begin{itemize}
  \item $X_{1t}, \ldots, X_{kt}$ are the exogenous climate variables at time $t$ (e.g., precipitation, average temperature),
  \item $\beta_1, \ldots, \beta_k$ are their corresponding coefficients.
\end{itemize}
```
In parallel, a separate GARCH model will be fitted to model time-varying
volatility in weekly cocoa prices, which is a prominent feature in
commodity markets. The GARCH model is capable of accounting time-varying
conditional variance, making it well-suited for capturing market
dynamics characterized by volatility clustering.

## GARCH Model

The GARCH(1,1) model is used to capture volatility in weekly cocoa price
returns. It helps estimate how uncertainty changes over time, offering
insight into risk and price fluctuations beyond average trends.

```{=latex}
\[
\begin{aligned}
r_t &= \mu + \varepsilon_t, \\
\varepsilon_t &= \sigma_t z_t, \\
\sigma_t^2 &= \omega + \alpha_1 \varepsilon_{t-1}^2 + \beta_1 \sigma_{t-1}^2
\end{aligned}
\]

where:

\begin{itemize}
  \item $r_t$ is the log return of weekly cocoa price at time $t$,
  \item $\sigma_t^2$ is the time-varying conditional variance (volatility),
  \item $z_t$ is an i.i.d. innovation term, assumed to follow a Student's $t$-distribution,
  \item $\omega > 0$, $\alpha_1 \geq 0$, and $\beta_1 \geq 0$ are parameters to be estimated.
\end{itemize}
```
Model orders for ARIMA and ARIMAX were selected using AIC, BIC, and
ACF/PACF diagnostics to balance fit and complexity. GARCH(1,1) was
chosen for its simplicity and effectiveness in capturing volatility
clustering, with diagnostics confirming model adequacy.

## VARMA Model

```{=tex}
\begin{align*} 
\text{PriceDiff}_{t} = &\ \beta_0 + \beta_1 \text{PriceDiff}_{t-1} + \beta_2 \text{ExRateDiff}_{t-1} + \beta_3 \text{MonthlyAvgTemp}_{t-1} + \\
&\ \beta_4 \epsilon_{\text{PriceDiff}, t-1} + \beta_5 \epsilon_{\text{ExRateDiff}, t-1} + \beta_6 \epsilon_{\text{MonthlyAvgTemp}, t-1} + \\
&\ \beta_7 \epsilon_{\text{PriceDiff}, t-2} + \beta_8 \epsilon_{\text{ExRateDiff}, t-2} + \beta_9 \epsilon_{\text{MonthlyAvgTemp}, t-2} + \epsilon_t
\end{align*}
```
Where:

-   `p = 1` means the model uses the lagged values from the previous
    time step (t-1) for PriceDiff, ExRateDiff, and MonthlyAvgTemp (AR
    part)

-   `q = 2`: means the model uses lagged residuals (errors) from the
    previous two time steps (t-1 and t-2) for PriceDiff, ExRateDiff, and
    MonthlyAvgTemp (MA part).

-   $\beta_1, \beta_2, ...$ are the coefficients for the lagged
    variables and error terms $\epsilon_t$ represents the residual
    (error) term at time t.

For modeling multivariate interdependencies, the VARMA model will be
used to fit the monthly cocoa price, including the exchange rate as
exogenous variables and changing the weekly precipitation and
temperature into monthly frequency. The cocoa price series will be
analyzed alongside relevant exogenous variables with each tested for
stationarity using the ADF test, followed by differencing where
necessary. The optimal lag structure for the VARMA model was identified
using the VARselect function from the “var” package, where various VARMA
models with the selected lags will be fitted and the final model will be
determined by having the lowest AIC. The fitted VARMA model allowed for
joint modeling of endogenous and exogenous variables, capturing the
dynamic interrelationship among them.

## XGBoost

To address nonlinearities and improve model robustness, XGBoost was also applied 
as a machine learning forecasting approach and it is selected for its ability to 
manage both linear and nonlinear relationships and its flexibility in dealing 
with non-stationary data without requiring transformation assumptions. 
Feature selection was conducted based on relative gain values, with variables 
above a 0.1 threshold retained to enhance interpretability and predictive 
accuracy. The XGBoost model will forecast the monthly cocoa price difference as a sum of multiple decision trees:

$$
\hat{y}_t = f_1(x_t) + f_2(x_t) + \dots + f_{500}(x_t)
$$

where $\hat{y}_t$ is the predicted **cocoa price difference** at time
$t$, $x_t$ represents the **input features** and $f_m(x_t)$ is the
**m-th decision tree** trained by XGBoost with total number of 500 boosting 
iterations.

# Data
## Weekly Data Description for ARIMA, ARIMAX, and GARCH Analysis

The data set used in this analysis consists of weekly observations on
international cocoa prices and climate indicators from Ghana.

-   **Cocoa Price Data**: Daily price series from the International
    Cocoa Organization (ICCO), provided in USD per metric tonne. This
    data set was aggregated to a weekly frequency by selecting the last
    available daily price for each week.

-   **Ghana Climate Data**: Daily weather data for multiple Ghanaian
    weather stations obtained from the National Centers for
    Environmental Information (NCEI). Variables include: Total daily
    precipitation (PRCP), Average daily temperature (TAVG), Maximum
    temperature (TMAX), and Minimum temperature (TMIN)

These were cleaned, averaged across stations, and aggregated to weekly
summaries (sum for precipitation, average for temperatures). Missing
values were forward and backward filled using Last Observation Carried
Forward (LOCF).

The final weekly data set spans from **January 2000 to December 2024**,
containing the following variables:

```{r weekly-summary-stats, echo=FALSE, message=FALSE, results='asis'}
summary_table <- data.frame(
  Variable = c("Cocoa Price (USD/tonne)", 
               "Total Weekly Precipitation (mm)", 
               "Average Weekly Temperature (°C)", 
               "Max Weekly Temperature (°C)", 
               "Min Weekly Temperature (°C)"),
  Min = c(774.1, 0.0, 74.0, 79.0, 66.6),
  `1st Qu.` = c(1557.99, 0.00, 78.99, 86.00, 72.85),
  Median = c(2192.96, 0.15, 81.32, 89.47, 74.00),
  Mean = c(2290.50, 0.44, 81.05, 88.83, 74.09),
  `3rd Qu.` = c(2715.88, 0.62, 83.00, 91.47, 75.34),
  Max = c(10966.97, 6.06, 88.14, 99.00, 80.29)
)

colnames(summary_table) <- c("Variable", "Min", "1st Qu.", "Median",
"Mean", "3rd Qu.", "Max")
#summary_table$Variable <- paste0("<span style='font-size: 10px;'>", summary_table$Variable, "</span>")
knitr::kable(summary_table, caption = "Summary Statistics for Weekly Cocoa Price and Climate Variables")
```

## Monthly Data Description for VARMA and XGBoost Analysis

The monthly dataset combines cocoa prices, climate indicators, exchange
rates, and production volumes to support multivariate modeling of
Ghana’s cocoa sector.

-   **Cocoa Price Data**: Sourced from the International Cocoa
    Organization (ICCO), daily prices in USD per tonne were aggregated
    to monthly frequency using the monthly average of daily values.

-   **Ghana Climate Data**: Ghana Climate Data: Daily weather
    observations (PRCP, TAVG, TMAX, TMIN) from multiple stations were
    aggregated using the monthly mean across all available station
    values.

-   **Exchange Rate Data**: Monthly average USD to GHS exchange rates
    were sourced from Investing.com [@USDGHSEntry]. These rates reflect the
    macroeconomic environment relevant to Ghana's cocoa exports.

-   **Cocoa Production Data**: Annual cocoa production volumes for Ghana
    were obtained from Our World in Data [@CocoaBeanProduction]. Production is reported by crop
    year (October–September), and annual totals were evenly distributed
    across months. This method is a practical simplification due to the
    lack of higher-frequency production data.

The final monthly dataset spans from **March 1995 to November 2024**,
and includes variables for cocoa prices, climate conditions, exchange
rates, and estimated production.

```{r monthly-summary-stats, echo=FALSE, message=FALSE, results='asis'}
summary_table_monthly <- data.frame(
  Variable = c("Cocoa Price (USD/tonne)",
           	"Monthly Average Temperature (°C)",
           	"Monthly Average Precipitation (mm)",
           	"Cocoa Production (Tonnes)",
           	"Exchange Rate (USD/GHS)"),
  Min = c(800.9, 76.27, 0.00000, 288075, 0.0995),
  `1st Qu.` = c(1567.9, 78.93, 0.01570, 436600, 0.8047),
  Median = c(2205.3, 81.42, 0.05727, 700020, 1.4318),
  Mean = c(2295.5, 81.09, 0.07431, 662729, 2.9517),
  `3rd Qu.` = c(2684.8, 82.83, 0.10840, 835466, 4.3690),
  Max = c(9876.6, 86.90, 0.45694, 1047385, 16.2500)
)

colnames(summary_table_monthly) <- c("Variable", "Min", "1st Qu.", "Median",
                                 	"Mean", "3rd Qu.", "Max")

knitr::kable(
  summary_table_monthly,
  label = "tab:monthly_summary_stats",
  caption = "Summary Statistics for Monthly Cocoa Price, Climate, Production, and Exchange Rate Variables",
  booktabs = TRUE,
  align = "lccc"
)
```

We now consider the VARMA model due to the volatility of cocoa prices,
while still incorporating parameters from the ARMA model. To ensure the
variables are strong significant (p-value \< 0.05), we first apply set
difference on the variables that are not statistically significant
(p-value \> 0.05), such as `cocoa_price`, `price_lag`, and
`currency_exchange_rate`. With these significant variables identified,
we proceed to fit the VARMA model by selecting an appropriate lag order.

```{r}
#| echo: false

lag_selection_table <- data.frame(
  Criterion = c("AIC(n)", "HQ(n)", "SC(n)", "FPE(n)"),
  `Selected Lag` = c(4, 2, 1, 4)
)
knitr::kable(lag_selection_table, caption = "Lag Order Selection")
```

From the results, we will consider lag=1, 2, 4. We will use model
selection to decide which one is better.

```{r}
#| echo: false

library(knitr)

aic_table <- data.frame(
  Model = c("VARMA(p=1, q=2)", "VARMA(p=2, q=2)", "VARMA(p=4, q=2)"),
  AIC = c(10677.67, 10713.37, 10683.50)
)
knitr::kable(aic_table, caption = "AIC for Different Lag Orders")
```

We built three VARMA models with different lag orders (p = 1, 2, 4, q =
2) to forecast price fluctuations. The difference between these models
lies in the number of past observations (lags) they consider for
predicting future values. To compare them, we used the AIC, which
balances model fit and complexity. The results show that VARMA(p=1, q=2)
has the lowest AIC, making it the best model among the three. We will
continue the model check with VARMA(1, 2) model.

![The residuals in the VARMA(1, 2) model shows that most of the bars,
which represent autocorrelation at different lags, fall within the blue
confidence interval bands, typically the default for ACF plots. This
indicates that the residuals do not exhibit significant autocorrelation
and are random. Therefore, the residuals can be considered white noise,
suggesting that the model has effectively captured the underlying
patterns. Based on this, we will forecast the price fluctuations over
time using VARMA(1, 2)
model.](~/STA457-Assignments/Cocoa_price_forecast/script/draft/VARMA-model-check.png){#fig:garch-volatility}

In XGBoost Model, the initial model will incorporate the following predictor variables
(features):

```{=tex}
\begin{itemize}
    \item $\text{Monthly\_Avg\_Temp}$ (Monthly Average Temperature)
    \item $\text{Monthly\_Avg\_Precip}$ (Monthly Average Precipitation)
    \item $\text{Exchange\_Rate}_t$ (Exchange rate)
    \item $\text{Production\_in\_Ton}_t$ (Cocoa Production in tons)
    \item $\text{Price\_diff}_{t}$ (Price difference)
    \item $\text{Price\_diff}_{t-1}$ (Price difference with 1 lag)
    \item $\text{Price\_diff}_{t-2}$ (Price difference with 2 lag)
    \item $\text{Price\_diff}_{t-3}$ (Price difference with 3 lag)
\end{itemize}
```
The inclusion of 1-month lagged exchange rate follows Frankel and
Hardouvelis' [@Frankel1985] observation supports that modeling exchange
rate effects with 1-month delay account for market absorption time for
currency fluctuations and reducing noise compared to immediate-rate
models. Furthermore, three lagged price differences is grounded in both
empirical evidence and commodity market dynamics, particular in
agricultural commodities like cocoa, where short-term price momentum and
cyclical patterns are exhibited frequently due to supply chain delays
coming from production adjustments and inventory restocking typically
manifest over quarterly horizons [@Gilbert2010].

\newpage

# Forecasting and Results
The performance of classical time series models like ARIMA and ARIMAX was assessed using weekly cocoa price data. Although these models are useful for understanding temporal dependencies and the impact of climate variables, their predictive performance was limited. Forecast accuracy metrics (MAE, RMSE, MAPE) indicated relatively high values for all ARIMA and ARIMAX models, with MAE ranging from 5009 to 5052, RMSE from 5429 to 5470, and MAPE from 175.41% to 176.77%, suggesting a poor predictive fit. This underperformance highlights the limitations of ARIMA in capturing cocoa price dynamics. In contrast, models like VARMA and XGBoost showed improved forecast accuracy, indicating a better ability to model the underlying data patterns. The VARMA model showed an RMSE of 1331.574 and an MAE of 1097.820, indicating a relatively higher magnitude of forecast errors compared to XGBoost. A GARCH(1,1) model was also applied to log returns to capture volatility in weekly cocoa prices. The model showed strong evidence of conditional heteroskedasticity and high volatility persistence (α₁ + β₁ ≈ 0.99), consistent with commodity market behavior. Residual diagnostics confirmed the model’s adequacy. While GARCH does not enhance point forecast accuracy, it effectively models the uncertainty and risk in price movements. In comparison, XGBoost, with MAE of 38.11, RMSE of 235.46, and MAPE of 41.14%, outperformed the other models, providing the most accurate forecast for cocoa prices.

```{r model-metrics-long-format, echo=FALSE, message=FALSE, results='asis'}
library(knitr)
# Create the transposed data frame
model_metrics <- data.frame(
  Model = c("ARIMAX(2,1,2)", "ARIMAX(1,1,1)", "ARIMA(0,1,1)", "ARIMA(0,1,2)",
        	"VARMA", "XGBoost"),
  MAE = c(5018.82, 5009.09, 5034.85, 5052.54, 1097.82, 38.11),
  RMSE = c(5438.84, 5429.23, 5452.90, 5470.55, 1331.57, 235.46),
  MAPE = c(175.73, 175.41, 176.20, 176.77, 97.87, 41.14)
)
# Display with kable
kable(
  model_metrics,
  caption = "Forecast Error Metrics by Model. ARIMA and ARIMAX models are based on weekly data, while VARMA and XGBoost use monthly data. Lower values for MAE, RMSE, and MAPE indicate better forecast accuracy.",
  booktabs = TRUE,
  align = "lccc"
)
```

\newpage

![Weekly cocoa log returns with GARCH(1,1) estimated volatility. The
dashed line shows changing volatility over time, highlighting periods of
greater market
risk.](~/STA457-Assignments/Cocoa_price_forecast/script/draft/ARIMA_GARCH_Volatility.png)

\newpage

![The forecast for price changes indicates significant short-term
volatility, with an initial sharp increase followed by fluctuations
before stabilizing around a moderate positive value. The first two steps
show a large jump (+88.74 to +241.41), followed by a drop (-20.06), and
then a gradual stabilization around 20–30. This suggests that prices may
experience a sharp initial surge before settling into a more predictable
upward trend. If applied to actual prices, this pattern implies a
high-risk, high-reward scenario in the short term, followed by steady
growth. For trading or investment decisions, this means navigating an
initial period of uncertainty before benefiting from more stable price
movements.](~/STA457-Assignments/Cocoa_price_forecast/script/draft/VARMA-Forecast.png)

![Feature Importance of Three Models: As XGBoost ranks feature by its
importance and highlights which variables drive price fluctuations,
focusing on substantial predictive importance features by dropping
variables with importance below 0.1, the final XGBoost model is
determines by the features Exchange rate, Price difference with 1 lag,
Price difference with 2
lag](~/STA457-Assignments/Cocoa_price_forecast/script/XGBoost_Draft/XGBoost-features.png)

![The discrepancies between the actual and predicted prices over the
test period from January 1, 2010, to November 1, 2024 with the price
differences fluctuate with notable deviations indicates the model
significantly overestimated the actual prices. The trend over time shows
variability with no clear consistent pattern, suggesting that the
model's performance may be influenced by external factors or inherent
limitations in capturing the underlying price
dynamics.](~/STA457-Assignments/Cocoa_price_forecast/script/XGBoost_Draft/XGBoost_ActualVSDiff.png){#fig:actual-vs-predict}

![Since XGBoost predicts price differences $\hat{y}_t$, the actual cocoa
price is calculated recursively by
$\text{Price}_t = \text{Price}_{t-1} + \hat{y}_t$, where price at time
$t$ is derived by adding the predicted difference to the previous price.
According to model 3, the forecasted monthly cocoa price during the
period from December 2024 to May 2025 is increasing linearly
approximately from \$8000 to
\$9800](~/STA457-Assignments/Cocoa_price_forecast/script/XGBoost_Draft/XGBoost_6MonthForecasting.png){#fig:forecast}

\newpage

# Discussion and Conclusion
## Discussion
This analysis applied ARIMA, ARIMAX, VARMA, and XGBoost models to forecast future cocoa prices and assess volatility with GARCH. Traditional time series models provided a useful baseline and offered interpretability, but their forecasting performance was limited. The GARCH(1,1) model effectively captured volatility clustering, reflecting persistent risk in cocoa price movements. ARIMAX models included climate-related variables (such as precipitation and temperature), but these exogenous variables had limited influence on forecasting accuracy. Their coefficients were small and often not statistically significant, suggesting that while they may impact cocoa production over the long term, they did not contribute meaningfully to short-term price forecasts in this setting. XGBoost significantly outperformed all other models, including VARMA(1, 2), demonstrating a strong ability to capture nonlinear and complex relationships. This suggests that hybrid approaches may offer improved accuracy and robustness in forecasting commodity prices.

## Conclusion
In our analysis of various forecasting models for cocoa prices, we evaluated several approaches, including ARIMA, ARIMAX, GARCH, VARMA, and XGBoost, using key performance metrics like MAE, RMSE, and MAPE. Among these models, XGBoost emerged as the top performer, delivering the most accurate predictions with the lowest error rates. The ARIMA and ARIMAX models, while useful in capturing price movements, demonstrated relatively higher forecast errors, especially in terms of MAE and RMSE. The GARCH model, although effective in modeling volatility and uncertainty, did not significantly improve the point forecast accuracy, highlighting its strength in risk management rather than direct price forecasting. VARMA, on the other hand, showed very poor performance, with notably high errors across all metrics, making it less suitable for this task.

After comparing the models, we chose XGBoost due to its superior ability to capture the underlying patterns in the cocoa price data, offering the most precise forecast while minimizing prediction error. As a result, the XGBoost model forecasts a steady increase in the monthly cocoa price from December 2024 to May 2025, rising from approximately \$8000 to \$9800. This forecast reflects a strong upward trend, likely driven by market factors such as supply-demand dynamics and broader economic influences.

## Limitations
Forecasting Ghana’s cocoa prices involved several limitations that potentially affected model accuracy and generalizability. A key challenge was data availability and quality. Crucial economic indicators, such as global demand and supply indices, were either inaccessible or available only upon request, limiting the inclusion of important explanatory variables. The dataset from Ghana’s Cocoa Board also lacked full temporal coverage. Additionally, daily climate data contained substantial missing values and spatial inconsistencies across weather stations further reduced reliability. Methodologically, reliance on univariate time series models (e.g., ARIMA) constrained the analysis of external influences. Although multivariate models like VARMA offer a more comprehensive approach, their implementation was hindered by data sparsity and complexity. In the XGBoost model, several features had to be excluded due to excessive missingness, potentially reducing predictive power. Lastly, access to advanced tools and datasets was limited by paywalls and proprietary restrictions, impeding broader model exploration.

## Further Suggestion
To enhance the robustness and accuracy of future forecasting efforts, several directions are worth considering. First, the integration of deep learning models, such as Long Short-Term Memory (LSTM) networks or Transformer-based architectures, may offer significant advantages in capturing complex, nonlinear patterns and long-range dependencies inherent in volatile cocoa markets. These models have shown promise in similar agricultural forecasting tasks and may outperform traditional and machine learning approaches, particularly when sufficient data is available.

Second, future research would benefit from incorporating higher-frequency data, particularly daily indicators of supply and demand dynamics, rather than relying solely on aggregate annual production figures. This would enable a more responsive modeling framework, better suited to capturing short-term market shifts driven by logistical, economic, or geopolitical events.

Finally, it is recommended that researchers explore the possibility of direct engagement with relevant institutions, such as governmental agencies, commodity boards, or international trade organizations. Formally requesting access to proprietary datasets may yield more granular and timely information, thus supporting the development of richer and more accurate forecasting models.

\newpage

# References
